---
title: "Estimating and filtering out ambient mRNA for sc-/sn-RNA-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  root_dir: './'
  metadata_dir: './'
  metadata_file: '.'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
  data_dir: './'
  soup_fraction_value_default: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```


\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak

# Information about this notebook
[SoupX method](https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html) profiles “the soup”, i.e., collection of cell-free mRNAs floating in the input solution. The soup looks different for each input solution and strongly resembles the expression pattern obtained by summing all the individual cells ([Young and Behjati, 2020](https://doi.org/10.1093/gigascience/giaa151)).

SoupX calculates `Cell-specific contamination fraction` (Estimate (or manually set) the contamination fraction, the fraction of UMIs originating from the background, in each cell) and infers a `corrected expression matrix` (Correct the expression of each cell using the ambient mRNA expression profile and estimated contamination).

The method to do this consists of three parts:

1. Calculate the profile of the soup.
2. Estimate the cell specific contamination fraction.
3. Infer a corrected expression matrix.


# Set up

We will set up libraries necessary for this notebook.

```{r load-library, echo=TRUE}
suppressPackageStartupMessages({
  library(future)
  library(knitr)
  library(SoupX)
  library(Seurat)
  library(stringr)
  library(tidyverse)
  library(tinytex)
  #library(SingleCellExperiment)
  library(Matrix)

  # Evaluate Seurat R expressions asynchronously when possible using future package
  source(normalizePath(file.path("..", "..", "scripts", "snap_parallel_plan.R"), winslash = "/"))
  snap_set_future_plan()})
```

# Directories and paths to file Inputs/Outputs

We will set up directories and paths necessary for this notebook.


```{r set-dir-and-file-names, echo=TRUE}
attach(params)

analysis_dir <- file.path(root_dir, "analyses", "upstream-analysis") 
module_results_dir <- file.path(analysis_dir, paste0("results"))

# Input files
project_metadata_file <- file.path(metadata_dir, metadata_file) # metadata input file

# Create SoupX_results_dir
SoupX_results_dir <- file.path(module_results_dir, paste0("01_SoupX"))
if (!dir.exists(SoupX_results_dir)) {
  dir.create(SoupX_results_dir)}

# Create to plots directory
plots_dir <- file.path(analysis_dir, "plots") 

# Create SoupX_plots_dir
SoupX_plots_dir <- file.path(plots_dir, paste0("01_SoupX"))
if (!dir.exists(SoupX_plots_dir)) {
  dir.create(SoupX_plots_dir)}

source(paste0(root_dir, "/figures/scripts/theme_plot.R"))
```

```{r echo=FALSE,warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H',
               fig.path = paste0(SoupX_plots_dir, "/",  "SoupX-contamination-fraction-plot-", sep = ""))
a4width <- 8.3
a4height <- 11.7
```

# Read metadata file and define `sample_name`
```{r read-metadata-define-sample-name, echo=TRUE}
# Read metadata
project.metadata <- read.csv(project_metadata_file, sep = '\t', header = TRUE)
sample_name <- unique(as.character(project.metadata$ID))
sample_name <- sort(sample_name, decreasing = FALSE)
print(sample_name)
```

# Read in 10x data and run SoupX

We will estimate the contamination fraction `rho`. The posterior distribution is calculated using a Poisson likelihood with a gamma distribution prior, parametrised by its mean priorRho and standard deviation priorRhoStdDev. The dotted line in the above plot shows the prior distribution. The default parameters have been calibrated to be fairly non-specific with a slight preference towards values of rho in the 0% to 10% range which is most commonly seen for fresh (i.e. not nuclear) single cell experiments. The default values place only a very weak constraint, as can be seen by setting a uniform prior. 

If contamination fraction estimate is > 20%, it is recommended to set `rho` manually. In this case, we will set 5% background to be removed by default.

```{r run-SoupX, echo=TRUE}
# Create contamination summary table
contamination_summary <- matrix(ncol = 4, nrow = 1) %>% as.data.frame()
colnames(contamination_summary) <- c("sample_name", "rho_estimated_by_SoupX", "rho_adjusted", "rho_removed")

for (i in seq_along(sample_name)) {
  
  cat("Beginning to process sample", sample_name[i], "\n")

  # Create directory 
  samples_SoupX_results_dir <- file.path(SoupX_results_dir, sample_name[i]) 
  if (!dir.exists(samples_SoupX_results_dir)) {
    dir.create(samples_SoupX_results_dir)}
  print(samples_SoupX_results_dir)

  #strainedCounts_SoupX_results_dir <- file.path(samples_SoupX_results_dir, "strainedCounts") 
  #if (!dir.exists(strainedCounts_SoupX_results_dir)) {
  #  dir.create(strainedCounts_SoupX_results_dir)}
  # print(strainedCounts_SoupX_results_dir)
  
  cat("Reading in data\n")
  
  filtered_matrix_file <- file.path(data_dir, sample_name[i], "outs", "filtered_feature_bc_matrix.h5") 
  raw_matrix_file <- file.path(data_dir, sample_name[i], "outs", "raw_feature_bc_matrix.h5") 
  clusters_file <- file.path(data_dir, sample_name[i], "outs", "analysis", "clustering", "gene_expression_graphclust", "clusters.csv") 

  # Read data
  filtered_matrix <- Seurat::Read10X_h5(filtered_matrix_file, use.names = TRUE)
  raw_matrix <- Seurat::Read10X_h5(raw_matrix_file, use.names = TRUE)
  clusters <- read.csv(clusters_file)
  
  # Save plot
  #pdf(file = paste0(SoupX_plots_dir, "/", sample_name[i], "_SoupX_contamination_fraction_plot.pdf", sep = ""))
  
  # Create df
  mDat <- data.frame(clusters=clusters$Cluster, row.names=clusters$Barcode)
  
  # Make a “SoupChannel”, the object needed to run SoupX
  sc_raw <- SoupChannel(raw_matrix, filtered_matrix, mDat)

  # Estimate contamination fraction rho
  sc <- autoEstCont(sc_raw, forceAccept = TRUE)
  print(sc)
  soup_fraction_value <- sc$fit$rhoEst
  
  # To not remove more than 20% background
  if (soup_fraction_value > 0.2) {
    print(soup_fraction_value)
    cat("Estimated soup fraction >20%, set to 5% by default.\n")
    sc = setContaminationFraction(sc_raw, soup_fraction_value_default)
    contamination_summary[i,] <- c(sample_name[i], soup_fraction_value, "yes", soup_fraction_value_default)
    print(sc)
    } else {
      cat("Estimated soup fraction\n is ", `soup_fraction_value`, "\n")
      contamination_summary[i,] <- c(sample_name[i], soup_fraction_value, "no", soup_fraction_value)}
   
  # Correcting expression profile and clean data
  # This will contain a corrected matrix to be used in place of the original table of counts in downstream analyses.
  out <- adjustCounts(sc, roundToInt = TRUE) # roundToInt option to make sure we output integer matrix
  colnames(out) <- paste0(sample_name[i], ":", colnames(out))
  colnames(out) = gsub('-1$','', colnames(out)) # remove -1
  
  #DropletUtils:::write10xCounts(paste(SoupX_results_dir, "./strainedCounts", sep = ""), out_merge, overwrite = TRUE) #library(DropletUtils)
  #write10xCounts(path = file.path(SoupX_results_dir, sample_name[i], "strainedCounts"), x = out) #library(SingleCellExperiment)
  #Write10X(out_merge, path = file.path(SoupX_results_dir, "strainedCounts")) #library(Seurat)

  #writeMM(out, file = file.path(strainedCounts_SoupX_results_dir, "matrix.mtx"))
  
  # Write barcodes (column names of the count matrix)
  #write.table(colnames(out), file = file.path(strainedCounts_SoupX_results_dir, "barcodes.tsv"), 
  #          quote = FALSE, row.names = FALSE, col.names = FALSE)

  # Write features (row names of the count matrix)
  #write.table(rownames(out), file = file.path(strainedCounts_SoupX_results_dir, "features.tsv"),
  #          quote = FALSE, row.names = FALSE, col.names = FALSE)

  saveRDS(out, file.path(samples_SoupX_results_dir, "mtx_corrected_SoupX.rds")) # Save combined matrices
  
  cat("Processing complete for sample", sample_name[i], "\n")
  
  # Let's combine corrected matrices from all samples 
  # We will use this file in later stage of the `upstream-analysis` module
  #  if(i==1) { 
  #    out_merge <- out
  #    } else {
  #      out_merge <- cbind(out_merge, out) }
  #dev.off()
}
```

# Contamination fraction `rho` per sample
We will summarize the estimated and removed background due to contamination per sample.

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- contamination_summary
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
caption_value = "Summary of contamination fraction per sample"
print(knitr::kable(tables1, align = "lcccc", caption = caption_value))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

# Save output files

```{r save-output, echo = TRUE}
#saveRDS(out_merge, file.path(SoupX_results_dir, "SoupX_corrected_mtx_merged.rds")) # Save combined matrices

contamination_summary %>%
  write_tsv(file.path(SoupX_results_dir,"SoupX_contamination_summary.tsv")) # Save `contamination_summary`
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r}
sessionInfo()
```

