---
title: "Doublet detection for sc-/sn-RNA-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  root_dir: './'
  metadata_dir: './'
  metadata_file: '.'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---
```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```


\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook
This script used the [ScDblFinder](https://bioconductor.org/packages/devel/bioc/vignettes/scDblFinder/inst/doc/scDblFinder.html) method for identifying doublets/multiplets in single-cell data. The scDblFinder method combines the strengths of various doublet detection approaches, training an iterative classifier on the neighborhood of real cells and artificial doublets.

scDblFinder() has two main modes of operation: cluster-based or not. Both perform quite well (see [Germain et al., 2021](https://f1000research.com/articles/10-979)). In general, it is recommended the cluster-based approach in datasets with a very clear cluster structure, and the random approach in more complex datasets.

The input data for scDblFinder can be either i) a count matrix (full or sparse), with genes/features as rows and cells/droplets as columns; or ii) an object of class SingleCellExperiment. In either case, the object should not contain empty drops, but should not otherwise have undergone very stringent filtering (which would bias the estimate of the doublet rate). 

# Set up
```{r load-library, echo=TRUE}
suppressPackageStartupMessages({
  library(scDblFinder)
  library(Seurat)
  library(scater)
  library(future)
  library(tidyverse)
  library(grid)
  library(knitr)

  # Evaluate Seurat R expressions asynchronously when possible using future package
  options(future.globals.maxSize = future_globals_value)
  source(normalizePath(file.path("..", "..", "scripts", "snap_parallel_plan.R"), winslash = "/"))
  snap_set_future_plan()})
```

```{r echo=FALSE,warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```

# Directories and paths to file Inputs/Outputs
```{r set-dir-and-file-names, echo=TRUE}
attach(params)
analysis_dir <- file.path(root_dir, "analyses", "upstream-analysis") 
results_dir <- file.path(analysis_dir, "results") 
seurat_results_dir <- file.path(analysis_dir, "results", "02_Seurat_qc") 

# Input files
seurat_results_file <- c(dir(path = seurat_results_dir,  pattern = "seurat_obj_raw.rds", full.names = TRUE, recursive = TRUE))
project_metadata_file <- file.path(metadata_dir, metadata_file) # metadata input file

# Create ScDblFinder_results_dir
ScDblFinder_results_dir <- file.path(paste0(results_dir, "/", "03_scDblFinder", "/"))
if (!dir.exists(ScDblFinder_results_dir)) {
  dir.create(ScDblFinder_results_dir)}

# Create to plots directory
plots_dir <- file.path(analysis_dir, "plots") 
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)}

# Create SoupX_plots_dir
scDblFinder_plots_dir <- file.path(plots_dir, paste0("03_scDblFinder"))
if (!dir.exists(scDblFinder_plots_dir)) {
  dir.create(scDblFinder_plots_dir)}

source(paste0(root_dir, "/figures/scripts/theme_plot.R"))
```

# Read metadata file and define `sample_name`

```{r read-metadata-define-sample-name, echo=TRUE}
# Read metadata
project.metadata <- read.csv(project_metadata_file, sep = '\t', header = TRUE)
sample_name <- unique(as.character(project.metadata$ID))
sample_name <- sort(sample_name, decreasing = FALSE)
print(sample_name)
```

# Read raw seurat_objs 
We will load the raw seurat object as generated in the `02A_seurat_alignment_qc.Rmd` step of the pipeline. We will run dimensionality reductions and convert to single cell experiments.

# Run scDblFinder 
Next, the `scDblFinder function` is used to identify cells that are doublets based on cluster-based generation of artificial doublets. We will save the cells that need to be filtered based on scDblFinder analysis in a separate table.

```{r run-recoverDoublets-plot-predictions, fig.width = 15, fig.height = 4, fig.fullwidth = TRUE, echo=TRUE}
# Create metrics per sample
doublets_metrics <- data.frame(matrix(ncol = 4, nrow = 1))
colnames(doublets_metrics) <- c("sample_name", "singlet", "doublet", "doublets_pct_library") # provide column names

# Create list for seurat and sce
sce_obj_list <- list()
seurat_obj_with_doublets_list <- list()

for (i in seq_along(sample_name)) {
  print(sample_name[i])
  
  ###############################################################
  cat("Reading in data for sample", sample_name[i], "\n")
  
  sce_obj_list[i] <- readRDS(seurat_results_file[i])
  print(sce_obj_list)
  
  set.seed(1234) # Make this reproducible as UMAP algorithm is not deterministic
  
  # Normalize, find variable features and scale data
  sce_obj_list[[i]] <- sce_obj_list[[i]] %>%
    NormalizeData() %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
    ScaleData(verbose = FALSE) 

  # Run dim_reductions
  sce_obj_list[[i]] <- RunPCA(sce_obj_list[[i]], features = VariableFeatures(object = sce_obj_list[[i]]), npcs=30) %>%
    RunUMAP(dims = 1:20, n.neighbors=20L, reduction.name = "umap", reduction.key = "UMAP_", verbose = FALSE)

  sce_obj_list[[i]] <- as.SingleCellExperiment(sce_obj_list[[i]])
  ###############################################################
  cat("Beginning to run scDblFinder for sample", sample_name[i], "\n")
  
  # Run scDblFinder
  # bp <- BiocParallel::MulticoreParam(4, RNGseed=1234)
  # MulticoreParam uses fork(), which can bus-error in Singularity on NFS mounts.
  bp <- BiocParallel::SerialParam(RNGseed = 1234)
  BiocParallel::bpstart(bp)
  
  set.seed(1234) # Make this reproducible 
  sce_obj_list[[i]] <- scDblFinder(sce_obj_list[[i]], clusters = TRUE, BPPARAM = bp)
  BiocParallel::bpstop(bp)
  
  # Estimate singlets and doublets in the library
  singlets_doublets_number <- print(table(sce_obj_list[[i]]$scDblFinder.class)) %>%
    as.data.frame() %>% 
    mutate(prediction = Var1,
           count = Freq) %>% 
    dplyr::select(-Var1, -Freq)

  # Estimate pct of doublets
  doublets_pct_library <- print(table(sce_obj_list[[i]]$scDblFinder.class)[2]/
                                  (table(sce_obj_list[[i]]$scDblFinder.class)[1] + 
                                     table(sce_obj_list[[i]]$scDblFinder.class)[2])*100)
  
  # Format decimal places to keep two digits
  doublets_pct_library <- format(round(doublets_pct_library, 2), nsmall = 2)
  ########################################################################################################
  cat("Creating metrics for sample", sample_name[i], "\n")  
  singlet <- as.numeric(singlets_doublets_number %>% 
                          dplyr::filter(prediction == "singlet") %>% 
                          dplyr::select(-prediction) %>% 
                          dplyr::mutate(singlet = count) %>% 
                          dplyr::select(-count))

  doublet <- as.numeric(singlets_doublets_number %>% 
                          dplyr::filter(prediction == "doublet") %>%
                          dplyr::select(-prediction) %>%
                          dplyr::mutate(doublet = count) %>%
                          dplyr::select(-count))

  doublets_metrics[i,] <- c(sample_name[i], singlet, doublet, doublets_pct_library)
  ########################################################################################################
  doublets <- sce_obj_list[[i]]@colData[,str_detect(colnames(sce_obj_list[[i]]@colData),"scDblFinder")] %>%
    as.data.frame() %>% 
    tibble::rownames_to_column("cell")
  
  # Cells that we need to filter from the final object
  doublets_to_filter <- subset(doublets, doublets$scDblFinder.class == "doublet") 
  
  # Convert sce list into seurat list
  seurat_obj_with_doublets_list[[i]] <- as.Seurat(sce_obj_list[[i]], 
                                        counts = "counts", # includes spliced and unspliced transcripts
                                        data = "logcounts") # derived from the counts matrixs
  
  ########################################################################################################
  # Plot the Doublets predictions on the UMAP
  cat("Plotting doublets predictions on the UMAP for sample:", sample_name[i], "\n")
  # merge all three plots within one grid (and visualize this)
  fname_png <- paste0(scDblFinder_plots_dir, "/", sample_name[i], "_Doublets_prediction.png")
  print(fname_png)
  gridExtra::grid.arrange(plot1 <- plotUMAP(sce_obj_list[[i]], colour_by = "scDblFinder.cluster", point_size = 0.1) + ggtitle("scDblFinder clusters"),
                          plot2 <- plotUMAP(sce_obj_list[[i]], colour_by = "scDblFinder.score", point_size = 0.1) + ggtitle("Doublets score"),
                          plot3 <- plotUMAP(sce_obj_list[[i]], colour_by = "scDblFinder.class", point_size = 0.1) + ggtitle("Doublets prediction"),
                          ncol = 3, nrow = 1, top = paste0(sample_name[i], "_Doublets_prediction"), padding = unit(1.5, "line"))
   
  g <- gridExtra::arrangeGrob(plot1, plot2, plot3, ncol = 3, nrow = 1,
                              top = paste0(sample_name[i], "_Doublets_prediction"),
                              padding = unit(1.5, "line")) #generates g
  ggsave(file = fname_png, g, width = 15, height = 4, device = "png")
  #######################################################################################
  # Save results for all libraries for cells to be filtered out
  if(i == 1){
    doublets_to_filter_all <- doublets_to_filter
    } else {
      doublets_to_filter_all <- rbind(doublets_to_filter_all, doublets_to_filter)}
  }

# Save table with doublets_metrics
write_tsv(doublets_metrics, paste0(ScDblFinder_results_dir, "/", "doublets_metrics.tsv")) 
```

# Doublets detection per sample
We will summarize the estimated and removed doublets per sample.

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- doublets_metrics
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
caption_value = "Summary of doublets detection per sample"
print(knitr::kable(tables1, align = "lcccc", caption = caption_value))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

# Create seurat object with all samples
We will create seurat objects and merge them into one. This will be used for filtering out the doublets.

```{r create-object, echo=TRUE}
# Merge all seurat objects in the list into one 
# The first parameter of merge should be a Seurat object, the second (y) can be one Seurat object or a list of several.
merged_seurat_obj <- merge(x = seurat_obj_with_doublets_list[[1]], y = seurat_obj_with_doublets_list[-1],
                           project = PROJECT_NAME)

# Remove any columns with NAs
merged_seurat_obj@meta.data <- merged_seurat_obj@meta.data %>%
  select_if(~ !any(is.na(.)))
print(DefaultAssay(merged_seurat_obj) <- "RNA") # Switch the default to RNA

# Add metadata
cat("Adding metadata to merged_seurat_obj\n")
sample.metadata <- rbind(project.metadata[i, ], project.metadata[i, ][rep(1, length(merged_seurat_obj$nCount_RNA)-1), ])
merged_seurat_obj@meta.data <- cbind(merged_seurat_obj@meta.data, sample.metadata)

# Cells_number_before_filter per sample_id
print(table(merged_seurat_obj@meta.data$orig.ident))

saveRDS(merged_seurat_obj, file = paste0(ScDblFinder_results_dir, "/", "merged_seurat_obj_with_doublets.rds")) # save object
```

```{r filter-object, echo=TRUE}
# Number of singlet vs doublet in seurat object
#print(table(merged_seurat_obj@meta.data$scDblFinder.class))

#split_obj <- SplitObject(merged_seurat_obj, split.by = "scDblFinder.class") # Split object based on scDblFinder prediction
#filter_obj <- subset(split_obj$singlet) # Keep cells with singlet
#Idents(filter_obj) <- "orig.ident" # Set cell identity classes

# Cells_number_after_filter per `sample_name`
#print(table(filter_obj@meta.data$orig.ident))

#saveRDS(filter_obj, file = paste0(ScDblFinder_results_dir, "/", "merged_seurat_obj_without_doublets.rds")) # save object
```

# Next steps

We have identified doublets from the libraries. 

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r}
sessionInfo()
```

